---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


### Loading and preprocessing the data
Load the data from activity.zip.
No extra processing of the data necessary to get started.

```{r Load data}
unzip ('activity.zip', overwrite = TRUE)
activity <- read.csv('activity.csv')

```


### What is mean total number of steps taken per day?
Histogram of number of steps a day.

```{r Mean total number of steps taken per day}
daily_activity <- aggregate(steps ~ date, activity, sum)
hist(daily_activity$steps, breaks = seq(0, 25000, 2500), 
     main = "Histogram of number of steps per day",
     xlab = "Steps")
day_mean <- mean(daily_activity$steps, na.rm = T)
day_median <- median(daily_activity$steps, na.rm = T)
```

The mean number of steps per days is **`r format(day_mean, digits = 1)`** and the median is **`r day_median`**

### What is the average daily activity pattern?
The plot below shows the average daily activity pattern via a time series plot of the 5-minute interval and the average number of steps taken across all days. 

```{r Average daily activity pattern}
interval_mean <- aggregate(steps ~ interval, activity, mean)
plot(interval_mean$interval, interval_mean$steps, type = "l", 
     main = "Average daily activity pattern",
     xlab = "Interval", 
     ylab = "mean(Steps)")
max_interval <- interval_mean[which.max(interval_mean$steps),1]

```

The interval with max number of steps is **`r max_interval`**. I.e. 8:35 in the morning.


### Inputing missing values
The original dataset is enriched by replacing the missing values with the corresponding mean value for that interval.
Decided to go with the actual mean value and not the integer part of the mean.

```{r Replace missing values}
activity_filled <- activity
activity_filled$steps <- ifelse(is.na(activity_filled$steps), 
                                interval_mean$steps, 
                                activity_filled$steps)
daily_activity_filled <- aggregate(steps ~ date, activity_filled, sum)

hist(daily_activity_filled$steps, breaks = seq(0, 25000, 2500), 
     main = "Histogram of number of steps per day with replaced missing values", 
     xlab="Steps")

missing_values <- sum(is.na(activity$steps))
days_missing_values <- length(unique(subset(activity, is.na(steps))$date))
day_mean <- mean(daily_activity_filled$steps)
day_median <- median(daily_activity_filled$steps)
```

The number of missing values where **`r missing_values`** among **`r days_missing_values`** days.    
With the missing values replaced, the mean number of steps per days is now **`r format(day_mean, digits = 1)`**, and the median is **`r format(day_median, digits = 1)`**.
The mean number is the same as before, but the median has moved slightly.  
The number of days with around mean (1076) steps is 8 more; the days with the missing entries.


### Are there differences in activity patterns between weekdays and weekends?
By spliting the activities by day type we can compare the activity patterns between weekday and weekends.

```{r "Difference in activity pattern", fig.height=10}
daytype <- c("weekend", "weekday", "weekday", "weekday", "weekday", "weekday", "weekend")
all_activity_week <- activity_filled
all_activity_week$daytype <- factor(daytype[as.POSIXlt(activity_filled$date)$wday + 1])
daily_activity_daytype <- aggregate(steps ~ daytype + interval, all_activity_week, mean)
weekdays <- subset(daily_activity_daytype, daytype == "weekday")
weekenddays <- subset(daily_activity_daytype, daytype == "weekend")

par(mfrow = c(2,1))
plot(weekdays$interval, weekdays$steps, type = "l", 
     main = "Weekdays",
     xlab = "Interval",
     ylab = "mean(Steps)")
plot(weekenddays$interval, weekenddays$steps, type = "l", 
     main = "Weekends",
     xlab = "Interval",
     ylab = "mean(Steps)")
```

The activities starts a bit later during the weekends. They are also spread out more during the day than on the weekdays.

#### Reference:
[Reproducible Resarch, Peer Assessment 1](https://class.coursera.org/repdata-034/human_grading/view/courses/975147/assessments/3/submissions)